\large \textbf{Burrows-Wheeler transform for terabases}

### 2015-10-22 paper/paper.tex
Multiple sections.

% Names
\newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
\newcommand{\sdarray}{\textsf{sdarray}}

@@

\begin{abstract}
Fooo
\end{abstract}

@@

\SubSection{Search}

Instead of searching for every text in collection $\Bcoll$ separately, we can search for the \emph{reverse trie} of the collection. Assume that there are $m_{\Acoll}$ texts in collection $\Acoll$ and $m_{\Bcoll}$ texts in collection $\Bcoll$. The \emph{root} of the trie corresponds to suffix $\$$, which has lexicographic rank $m_{\Acoll}$ in $\Acoll$ and corresponds to lexicographic range $[1,m_{\Bcoll}]$ in $\Bcoll$.

@@

\SubSection{Buffering}

To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk. Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}. The run buffer stores the runs as pairs of integers $(r, \ell)$. Once the run buffer becomes full, we sort the runs by \emph{run heads} $r$, use \emph{differential encoding} for the run heads, and encode the differences and run lengths with a \emph{prefix-free code}. The compressed run buffer is then merged with the similarly compressed thread buffer.

@@

\SubSection{Merge}

The merge phase starts with the rank array stored in multiple files on disk. Each of the files is sorted and compressed. For interleaving the \BWT{}s, we need to merge the files to produce an increasing run-length encoded stream of lexicographic ranks.

This work can also be parallelized between multiple threads. One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks. Another thread consumes the stream and uses it to interleave the \BWT{}s. Because the producer thread is usually slower than the consumer thread, we may want to use multiple threads for the multiway merge, if the disk is fast enough.

@@

The internal alphabet of \BWTmerge{} is \texttt{012345}, which corresponds to either \texttt{\$ACGTN} or \texttt{\$ACGNT}, depending on where the \BWT{}s come from. \BWT{}s using different alphabetic orders cannot be merged. We use simple byte-level codes for \emph{run-length encoding} the \BWT{}s. The encoding of run $(c, \ell)$, where $c$ is the character value and $\ell$ is the length, depends on the length of the run. If $\ell \le 41$, the run is encoded in a single byte as $6 \cdot (\ell-1) + c$. Longer runs start with byte $6 \cdot 41 + c$, followed by the encoding of $\ell-42$. The remaining run length is encoded as a sequence of bytes, with the low 7 bits containing data and the high bit telling whether the encoding continues in the next byte. The compressed buffers use the same 7+1\nobreakdash-bit code for both differentially encoded run heads and run lengths.

@@

environment, compiler

data: ReadServer BWTs, NA12878 trio \cite{1000GP2015}
error correction using bfc \cite{Li2015}

ReadServer: trade-offs with different parameters

trio: ropebwt + this vs ropebwt vs ropebwt2

>>>>

% Names
\newcommand{\BWTmerge}{\textsf{BWT\nobreakdash-merge}}
\newcommand{\sdarray}{\textsf{sdarray}}
\newcommand{\CEU}{\textsf{CEU}}
\newcommand{\RS}{\textsf{RS}}

@@

\begin{abstract}
In order to avoid the reference bias introduced by mapping reads to a reference genome, bioinformaticians are investigating reference-free methods for analyzing sequenced genomes. With large projects sequencing thousands of individuals, this raises the need for tools capable of processing terabases of sequence data. A key method is the Burrows-Wheeler transform (BWT), which is widely used for compressing and indexing reads. We propose a practical algorithm for building the BWT of a large read collection by merging the BWTs of smaller subcollections. The algorithm is capable of merging over 700 Gbp/day on a single system, with 20--25 gigabytes of memory overhead on top of the run-length encoded BWTs.
\end{abstract}

@@

\smallbreak\noindent\textbf{Search.} Instead of searching for every text in collection $\Bcoll$ separately, we can search for the \emph{reverse trie} of the collection. Assume that there are $m_{\Acoll}$ texts in collection $\Acoll$ and $m_{\Bcoll}$ texts in collection $\Bcoll$. The \emph{root} of the trie corresponds to suffix $\$$, which has lexicographic rank $m_{\Acoll}$ in $\Acoll$ and corresponds to lexicographic range $[1,m_{\Bcoll}]$ in $\Bcoll$.

@@

\smallbreak\noindent\textbf{Buffering.} To reduce disk usage and the amount of disk I/O, we buffer and compress the lexicographic ranks before writing them to disk. Each thread has two buffers: a \emph{run buffer} and a \emph{thread buffer}. The run buffer stores the runs as pairs of integers $(r, \ell)$. Once the run buffer becomes full, we sort the runs by \emph{run heads} $r$, use \emph{differential encoding} for the run heads, and encode the differences and run lengths with a \emph{prefix-free code}. The compressed run buffer is then merged with the similarly compressed thread buffer.

@@

\smallbreak\noindent\textbf{Merge.} The merge phase starts with the rank array stored in multiple files on disk. Each of the files is sorted and compressed. For interleaving the \BWT{}s, we need to merge the files to produce an increasing run-length encoded stream of lexicographic ranks. We can also use multiple threads. One thread reads the files and performs a \emph{multiway merge} using a priority queue, producing a stream of lexicographic ranks. Another thread consumes the stream and uses it to interleave the \BWT{}s. Because the producer thread is usually slower than the consumer thread, we may want to use multiple threads for the multiway merge, if the disk is fast enough.

@@

The internal alphabet of \BWTmerge{} is \texttt{012345}, which corresponds to either \texttt{\$ACGTN} or \texttt{\$ACGNT}, depending on where the \BWT{}s come from. \BWT{}s using different alphabetic orders cannot be merged. We use simple byte-level codes for \emph{run-length encoding} the \BWT{}s. The encoding of run $(c, \ell)$, where $c$ is the character value and $\ell$ is the length, depends on the length of the run. If $\ell \le 41$, the run is encoded in a single byte as $6 \cdot (\ell-1) + c$. Longer runs start with byte $6 \cdot 41 + c$, followed by the encoding of $\ell-42$. The remaining run length is encoded as a sequence of bytes, with the low 7 bits containing data and the high bit telling whether the encoding continues in the next byte. The compressed buffers use the same 7+1\nobreakdash-bit code for both the differentially encoded run heads and the run lengths.

@@

We ran our experiments on a system with two 16\nobreakdash-core AMD Opteron 6378 processors and 256 gigabytes of memory. The system was running Ubuntu 12.04 on Linux kernel 3.2.0. All software was compiled with gcc/g++ version 4.9.2.

\begin{table}[t!]
\begin{center}
\caption{Datasets used in the experiments.}\label{table:datasets}\smallskip%
{
\renewcommand{\baselinestretch}{1}\footnotesize
\begin{tabular}{cc|ccc}
\hline
\textbf{Dataset} & \textbf{Collection} & \textbf{Size} & \textbf{Reads} & \textbf{Read length} \\
\hline
\CEU             & NA12878             &       284 Gbp &          2.81G &               101 bp \\
\CEU             & NA12891             &       242 Gbp &          2.40G &               101 bp \\
\CEU             & NA12892             &       245 Gbp &          2.42G &               101 bp \\
\CEU             & Merged              &       771 Gbp &          7.63G &               101 bp \\
\hline
\RS              & AA                  &       433 Gbp &          4.69G &         73 or 100 bp \\
\RS              & TT                  &       432 Gbp &          4.68G &         73 or 100 bp \\
\RS              & AT                  &       275 Gbp &          2.98G &         73 or 100 bp \\
\RS              & TA                  &       355 Gbp &          3.84G &         73 or 100 bp \\
\RS              & Merged              &      1.49 Tbp &          16.2G &         73 or 100 bp \\
\hline
\end{tabular}}
\end{center}
\end{table}

We used two datasets from the phase 3 of the \emph{1000 Genomes Project} \cite{1000GP2015}:
\begin{itemize}

\item \CEU{} contains reads from high-coverage sequencing of the \emph{CEU trio} (individuals NA12878, NA12891, and NA12892). We downloaded the gzipped FASTQ files (run accessions SRR622457, SRR622458, and SRR622459) and concatenated the files representing the same individual. Then we corrected the sequencing errors with BFC \cite{Li2015} (\texttt{bfc -s 3g -t 16}) separately for each individual.

\item \RS{} is from the ReadServer project, which uses all sequence data from the phase 3. After error correction, trimming the reads to 73 or 100~bp, and merging the duplicates, there are 53.0~billion unique reads for a total of 4.88~Tbp. The reads are in 16 run-length encoded \BWT{}s built using the \emph{String Graph Assembler} \cite{Simpson2012}, distributed according to the last two bases. We chose four of the files (AA, TT, AT, and TA).

\end{itemize}
See Table~\ref{table:datasets} for further details on the datasets.

CEU: ropebwt + this vs ropebwt vs ropebwt2

RS: trade-offs with different parameters

Note: input/output formats, BWT-merge times include verification
